---
title: "HW03: Data Wrangling and Analysis of U.S. Supreme Court Decisions"
author: "Sophie Ennis"
output: github_document
---

## DATA

Before starting this homework, take some time to review the data description posted on our website under HW3. Familiarize yourself with each data source, the variables required for this assignment and their codebook. Understanding the data beforehand will save you time during the assignment and help prevent issues. Additionally, we encourage you to review the rubric used to evaluate homework assignments (see the website).


Load the libraries
```{r}
library(tidyverse)
library(readr)
```


Load the data using `read_csv()` and a relative path (notice the two data frames are located in the "data" folder). Label the imported data as `scdb_case` and `scdb_vote`
```{r}
scdb_case <- read_csv("data/scdb-case.csv")
scdb_vote <- read_csv("data/scdb-vote.csv")
```


Explore the data using functions such as `head`, `summary`, `glimpse`, etc. or any other suitable functions 
```{r}
glimpse(scdb_case)
glimpse(scdb_vote)
head(scdb_case)
head(scdb_vote)
```


## TASK 1: RECODE VARIABLES

Drawing from the content we have learned so far in the course, recode the following variables according to the instructions below. You will use the newly generated variables (along with other variables from the data) to answer the other parts of this assignment.

Important: 

* Pay attention to the data type: your recoded variable should have the same data type of the original variable, unless you are explicitly asked to modify the data type. For example, if the variable majority is a `dbl` (double), make sure that your recoded variable is also a `dbl`
* For each variable that you recode, show the total observations for each value before and after recoding it

Variables to recode:

`decisionDirection`: this variable takes four values (1 conservative, 2 liberal, 3 unspecifiable, and NA); please verify this in the [documentation](http://scdb.wustl.edu/documentation.php?var=decisionDirection#norms). Recode this variable to take three values: 0, 1, NA (0 conservative, 1 liberal, and NA for both NA and unspecifiable)
```{r}
scdb_case <- scdb_case %>% 
  mutate(decisionDirection_recode = case_when(
    decisionDirection == 1 ~ 0,
    decisionDirection == 2 ~ 1,
    decisionDirection == 3 ~ NA,
    is.na(decisionDirection) ~ NA))

print(scdb_case)

before_recode_dD <- scdb_case %>%
  count(decisionDirection, name = "total_observations_before")
after_recode_dD <- scdb_case %>%
  count(decisionDirection_recode, name = "total_observations_after")
comparison_dD <- left_join(before_recode_dD, after_recode_dD, 
                        by = c("decisionDirection" = "decisionDirection_recode"))
comparison_dD
```

`majority`: this variable takes two values (1 dissent, 2 majority; see [documentation](http://scdb.wustl.edu/documentation.php?var=majority). Recode it so dissent becomes 0 and majority becomes 1; leave the missing data unchanged
```{r}
scdb_vote <- scdb_vote %>% 
  mutate(majority_recode = case_when(
    majority == 1 ~ 0,
    majority == 2 ~ 1))
print(scdb_vote)

before_recode_m <- scdb_vote %>%
  count(majority, name = "total_observations_before")
after_recode_m <- scdb_vote %>%
  count(majority_recode, name = "total_observations_after")
comparison_m <- left_join(before_recode_m, after_recode_m, 
                        by = c("majority" = "majority_recode"))
comparison_m
```

`direction`: this variable takes two values (1 conservative, 2 liberal); see the [documentation](http://scdb.wustl.edu/documentation.php?var=direction). Recode it so conservative becomes 0 and liberal becomes 1; leave the missing data unchanged
```{r}
scdb_vote <- scdb_vote %>% 
  mutate(direction_recode = case_when(
    direction == 1 ~ 0,
    direction == 2 ~ 1))
print(scdb_vote)

before_recode_d <- scdb_vote %>%
  count(direction, name = "total_observations_before")
after_recode_d <- scdb_vote %>%
  count(direction_recode, name = "total_observations_after")
comparison_d <- left_join(before_recode_d, after_recode_d, 
                        by = c("direction" = "direction_recode"))
comparison_d
```

`chief`: convert this variable to a factor; check the original order of the levels of this variable (they should be Jay, Routlege, Ellsworth, etc.) and make sure your new factor variable has the same order
```{r}
chief_levels <- c("Jay", "Rutledge", "Ellsworth", "Marshall", "Taney", 
                  "Chase", "Waite", "Fuller", "White", "Taft", "Hughes", 
                  "Stone", "Vinson", "Warren", "Burger", "Rehnquist", 
                  "Roberts")
class(chief_levels)
scdb_case <- 
  scdb_case %>% 
  mutate(chief_factor = factor(chief, levels = chief_levels))
print(scdb_case)

before_recode_c <- scdb_case %>%
  count(chief, name = "total_observations_before")
after_recode_c <- scdb_case %>%
  count(chief_factor, name = "total_observations_after")
comparison_c <- left_join(before_recode_c, after_recode_c, 
                        by = c("chief" = "chief_factor"))
comparison_c
```

`justiceName`: convert this variable to a factor; your new factor variable should be in chronological order, meaning from the oldest to the newest judge in terms of time served in the Supreme Court; this order should be the same as in [`justice`](http://scdb.wustl.edu/documentation.php?var=justice)
```{r}
scdb_vote %>% distinct(justiceName) %>% pull(justiceName)
scdb_vote %>% distinct(justice) %>% print(n = 117)
```
```{r}
justiceName_levels <- c("JJay", "WCushing", "JWilson", "JBlair", "JIredell", 
                        "TJohnson", "WPaterson", "JRutledge2", "SChase", 
                        "OEllsworth", "BWashington", "AMoore", "JMarshall", 
                        "WJohnson", "HBLivingston", "TTodd", "GDuvall", "JStory", 
                        "SThompson", "RTrimble", "JMcLean", "HBaldwin", "JMWayne", 
                        "RBTaney", "PPBarbour", "JCatron", "JMcKinley", "PVDaniel", 
                        "SNelson", "LWoodbury", "RCGrier", "BRCurtis", "JACampbell", 
                        "NClifford", "NHSwayne", "SFMiller", "DDavis", "SJField", 
                        "SPChase", "WStrong", "JPBradley", "WHunt", "MRWaite", 
                        "JHarlan1", "WBWoods", "SMatthews", "HGray", "SBlatchford",
                        "LQLamar", "MWFuller", "DJBrewer", "HBBrown", "GShiras", 
                        "HEJackson", "EDEWhite", "RWPeckham", "JMcKenna", "OWHolmes", 
                        "WRDay", "WHMoody", "HHLurton", "CEHughes1", "WVanDevanter", 
                        "JRLamar", "MPitney", "JCMcReynolds", "LDBrandeis", "JHClarke", 
                        "WHTaft", "GSutherland", "PButler", "ETSanford", "HFStone", 
                        "CEHughes2", "OJRoberts", "BNCardozo", "HLBlack", "SFReed",
                        "FFrankfurter", "WODouglas", "FMurphy", "JFByrnes", "RHJackson",
                        "WBRutledge", "HHBurton", "FMVinson", "TCClark", "SMinton", 
                        "EWarren", "JHarlan2", "WJBrennan", "CEWhittaker", "PStewart", 
                        "BRWhite", "AJGoldberg", "AFortas", "TMarshall", "WEBurger",
                        "HABlackmun", "LFPowell", "WHRehnquist", "JPStevens", "SDOConnor",
                        "AScalia", "AMKennedy", "DHSouter", "CThomas", "RBGinsburg", 
                        "SGBreyer", "JGRoberts", "SAAlito", "SSotomayor", "EKagan", 
                        "NMGorsuch", "BMKavanaugh", "ACBarrett")
scdb_vote <- scdb_vote %>% 
  arrange(justice) %>% 
  mutate(justiceName_factor = factor(justiceName, levels = justiceName_levels))
print(scdb_vote)

before_recode_jN <- scdb_vote %>%
  count(justiceName, name = "total_observations_before")
after_recode_jN <- scdb_vote %>%
  count(justiceName_factor, name = "total_observations_after")
comparison_jN <- left_join(before_recode_jN, after_recode_jN, 
                        by = c("justiceName" = "justiceName_factor"))
comparison_jN
```


## TASK 2: ANSWER SPECIFIC QUESTIONS

The goal of this second task is to further explore the data and review fundamental `dplyr` commands. Each question in this task should be answered with just a few lines of code. Consult the [documentation](http://scdb.wustl.edu/documentation.php?s=1) to understand the meanings of the variables that you need to use.

Task 2 Question 1. Calculate the average number of minority (dissenting) votes cast per term using the appropriate variables (inspect the documentation to identify them). Exclude any missing values from your calculation using methods learned in the course. Provide a few sentences to explain the results (e.g., what does an average of 0.1 mean for a given year? what is the highest and lowest and what do they mean? etc.)
```{r}
dissent_per_term <- scdb_case %>% 
  group_by(term) %>% 
  summarize(avg_dissent = mean(minVotes, na.rm = TRUE))
print(dissent_per_term)
```
```{r}
summary(dissent_per_term)
```

An average of 0.1 for a given term means that the justices were largely in agreement during this term and most of the time did not dissent. The lowest average minority votes is 0.00, and the highest average is 2.15. This means that the highest number of dissenting votes that has ever been recorded is on average two votes per term, and the lowest is an average of zero dissenting votes per term. This means that the justices are often in agreement on their rulings.

Task 2 Question 2. Is chief Justice Earl Warren a conservative or liberal justice? To answer this question, compare how many times this justice voted in either direction. Use the the appropriate variables among those you recoded (check the documentation for details). Provide a few sentences to explain the results.
```{r}
combined_scdb <- scdb_case %>% 
  inner_join(scdb_vote, by = "caseId")

chief_warren <- combined_scdb %>% 
  filter(chief_factor == "Warren") %>% 
  group_by(direction_recode) %>% 
  summarize(vote_count = n())

print(chief_warren)
```

I recoded the variable direction so that conservative became 0 and liberal became 1. This tibble therefore shows that Chief Warren was a liberal justice, voting nearly twice as much liberally than he did conservatively.

Task 2 Question 3. Formulate a question of similar complexity to the preceding two questions and answer it using `dplyr`. The goal of this question is for you to explore the data as you prefer and reinforce your knowledge of fundamental `dplyr` verbs.

Which justice has voted the most conservatively?
```{r}
most_conservative <- combined_scdb %>% 
  group_by(justiceName_factor) %>% 
  summarize(conservative_count = 
              sum(direction_recode == 0, 
                  na.rm = TRUE)) %>% 
  arrange(desc(conservative_count))
print(most_conservative)
```

William Rehnquist has voted the most conservatively, voting 3,010 times conservatively. John Rutledge (#2) voted the least conservatively, with only 2 votes cast for conservative decisions.

## TASK 3: ANSWER OPEN-ENDED QUESTIONS

This final task builds on the previous two tasks. It simulates scenarios you might encounter in real research settings, requiring more lines of code to solve (e.g., remember to break down problems into simpler steps!) and a deeper understanding of the data, as well as of `dplyr` and `ggplot2`.

You can consult the `demo` folder in your repo for example plots. While we won’t accept solutions that replicate ours exactly, you may use them as a reference for your own work.

For each question, show graph(s) and provide 1-2 paragraphs of written interpretation of your results. The interpretation should be both descriptive (e.g., explaining what the graph depicts) and substantive (what do the results show, and your analysis and conclusions by observing the graph). Please add your written paragraphs as text (not as a R comments), and place them after the code and results.


Task 3 Question 1. What percentage of cases in each term are decided by a one-vote margin (i.e. 5-4, 4-3, etc.)? Show a graph with your results and interpret them.

In addition to the code with the complete solution and your interpretation of it, your submission should include the following two elements: code that shows your initial exploration of the variables needed to answer this question (e.g., if you need a categorical variable, use the most appropriate `dplyr` function(s) to display that variable categories and their counts, etc.); a brief explanation (as a text) of your logic/approach for calculating the percentage or proportion of cases in each term that are decided by a one-vote margin.

```{r}
glimpse(scdb_case)
summary_minVotes <- scdb_case %>% 
  summarize(minVotes_mean = mean(minVotes, na.rm = TRUE))
summary_minVotes

one_vote <- scdb_case %>% 
  group_by(term) %>% 
  summarize(total_cases = n(),
            one_vote = sum(minVotes == 4)) %>% 
            mutate(one_vote_percent = (one_vote / total_cases) * 100)
ggplot(one_vote,
       aes(x = term,
           y = one_vote_percent)) +
  geom_point() +
  labs(
    title = "Percentage of SCOTUS Cases Decided by a One Vote Margin",
    x = "Term",
    y = "Percentage of Cases") +
  theme_classic()
```
From 1800 to around 1850, there were very few cases where the Supreme Court disagreed on their decisions or was split by conservatives and liberals with very few minority votes cast, with clearly very few close margins. This starts to minimally change arond 1850, but the percentage of marginal minority votes that would create close calls stayed far below 10% until a bit before 1950. This is likely due to the Civil Rights movement where differences between conservatives and liberals were beginning to become more apparent in their effects on the Supreme Court. Thus, around 1950, the number of marginal cases decided by one vote exploded to 30% of cases, meaning the court was very split ideologically. While most cases from then on floated around 20% marginal, still many cases were decided by one vote well into the 2000s.

I used glimpse() to see the variables for scdb_case so I could make sure the minVotes variable was included in this dataset. I then calculated the mean number of minority votes per case and found it was less than 1, so the court must generally agree on most decisions. I wanted to use minVotes to calculate the cases decided by a one-vote margin because I knew that the decisions would be 5-4 because there are 9 justices and 4 would be the largest minority number of votes. I divided the number of cases that were decided with 4 justices voting for the minority opinion by the total number of cases to get the amount of cases that were decided by a one-vote margin, multiplying the decimal this calculated by 100 to get a percentage. I used a scatterplot because both term and the calculated percentage are continuous variables. 

Task 3 Question 2. How often the justices who are currently serving on the Supreme Court, have voted in the conservative direction in the following cases: criminal procedure, civil rights, economic activity, and federal taxation? Note these information might be stored in different dataframes
* Select only the relevant cases and select only the justices who are currently serving on the Supreme Court: John G. Roberts, Clarence Thomas, Samuel A. Alito, Sonia Sotomayor, Elena Kagan, Neil M. Gorsuch, Brett M. Kavanaugh, Amy Coney Barrett (Ketanji Brown Jackson is not in the data); see [here](https://www.supremecourt.gov/about/biographies.aspx) for details
* Calculate how often these justices have voted in the conservative direction in the selected cases
* Produce two graphs with your results: one faceted by justice and the other faceted by issue area, and interpret them (hint: to display % in the axes, explore the `ggplot2` documentation (e.g., `scale_y_continuous()` or `scale_x_continuous()`); to display the justice and issue area labels, you may want to convert them to factors first)

In addition to the code with the complete solution and your interpretation of it, your submission should include the following two elements: code that shows your initial exploration of the variables needed to answer this question (e.g., if you need a categorical variable, use the most appropriate `dplyr` function(s) to display that variable categories and their counts, etc.); a brief explanation (as a text) of your logic/approach for calculating how often the justices have voted in the conservative direction in the selected cases.


```{r}
current_justices <- c("CThomas", "JGRoberts", "SAAlito", 
                      "SSotomayor", "EKagan", "NMGorsuch", 
                      "BMKavanaugh", "ACBarrett")
relevant_issues <- c(1, 2, 8, 12)
relevant_cases <- scdb_case %>% 
  filter(issueArea %in% relevant_issues) %>% 
  inner_join(scdb_vote, by = "caseId") %>% 
  filter(justiceName %in% current_justices)
relevant_cases
```

```{r}
justice_counts <- relevant_cases %>%
  count(justiceName)
justice_counts

issue_counts <- relevant_cases %>%
  count(issueArea)
issue_counts
```

```{r}
relevant_cases <- relevant_cases %>% 
  mutate(direction_recode = case_when(
    direction == 1 ~ 0,
    direction == 2 ~ 1))
justice_conservative <- relevant_cases %>% 
  group_by(justiceName, issueArea) %>% 
  summarize(total_votes = n(),
            conservative_votes = sum(direction_recode == 0, na.rm = TRUE),
            conservative_pct = (conservative_votes / total_votes) * 100) %>% 
  ungroup()
justice_conservative
```

```{r}
ggplot(justice_conservative,
       aes(x = issueArea,
           y = conservative_pct,
           fill = justiceName)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "SCOTUS Conservative Decisions by Justice",
      x = "Issue Area",
      y = "Percentage of Conservative Votes") +
theme_classic() +
  facet_wrap(~ justiceName) +
  coord_flip()

ggplot(justice_conservative,
       aes(x = justiceName,
           y = conservative_pct,
           fill = issueArea)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "SCOTUS Conservative Decisions by Issue Area",
    x = "Justice",
    y = "Percentage of Conservative Votes") +
  theme_classic() +
  facet_wrap(~ issueArea) +
  coord_flip()
```
The justices that have voted the most conservatively overall based on the two graphs appear to be Alito, Thomas, Gorsuch, Barrett, and Kavanaugh. It is more clear when faceted by issue area that certain justices tend to vote more conservatively on different things, like how Barrett is the most likely out of all of the justices to vote conservatively on issues of federal taxation, coded as 12 in the codebook, while the other justices (besides Gorsuch) tend to vote more liberally on this issue. The issue of economic activity leaves all of the justices hovering around the 50% line for voting conservatively, meaning that they vote equally conservatively and liberally. On the issues of criminal procedure and civil rights, however, the differences between the liberal and conservative justices begin to become apparent, with Sotomayor and Kagan voting far less conservatively than the other justices. 

I chose to use the numeric representations of the issue areas and the abbreviated names of the justices because that is how they are present in the datasets. I utilized a similar coding strategy to the recoding of the direction variable and finding a percentage of votes from previous questions. I summed the conservative votes by summing the direction recoded variable when it was assigned 0, the conservative value. I calculated the percentage justices voted conservatively by dividing the summed conservative votes by each justice's total votes. I chose to use geom_col() because geom_bar() wasn't working, and used scale_y_continuous() to display percentages in the axes. I chose to fill and facet_wrap() the variable that was meant to be faceted and flipped the coordinates for easier readability.


Task 3 Question 3. Formulate a question of similar complexity to the preceding two questions and provide an answer that includes at least one graph. As in the previous two questions, include the initial code to explore the variables, a brief written explanation of your approach to solve the question, and provide 1-2 paragraphs with your interpretation of the results.
The goal of this question is for you to explore the data as you prefer and enhance your data analysis and coding skills.

```{r}
states_of_interest <- c(6, 12, 37, 51)

relevant_states <- scdb_case %>% 
  filter(petitionerState %in% states_of_interest | respondentState %in% states_of_interest)

relevant_states <- relevant_states %>%
  mutate(same_state = case_when(
    petitionerState == respondentState ~ "Same State",
    TRUE ~ "Different States"))
```

```{r}
relevant_states
petitioner_summary <- relevant_states %>%
  count(petitionerState)
petitioner_summary

respondent_summary <- relevant_states %>%
  count(respondentState)
respondent_summary
```

```{r}
state_counts <- relevant_states %>%
  group_by(petitionerState, respondentState, same_state) %>%
  summarize(total_cases = n(), .groups = 'drop')

same_diff_counts <- state_counts %>%
  group_by(same_state) %>%
  summarize(total = sum(total_cases), .groups = 'drop')
```

```{r}
ggplot(same_diff_counts, aes(x = same_state, y = total, fill = same_state)) +
  geom_bar(stat = "identity") +
  labs(title = "Cases by Same State vs. Different States",
       x = "Case Type",
       y = "Total Cases") +
  theme_classic()
```
Petitioners are the individuals that are suing the respondents, and I wanted to see if they were from the same state or not to potentially find evidence for different backgrounds contributing to the likelihood of suing each other. It seems like my hypothesis may be supported here, as almost 2000 cases have been pursued by people from different states, while only around 100 cases have been pursued by people from the same states. Of course, there is a lot of room for variation because of how many states there are, but I chose states with the largest populations in the country which should give the same states variable a better chance for people being from the same state, and still got very low case numbers for the same states variable. Being from different states may also exist along political lines which may influence different beliefs about civil rights, one of the issue areas brought to the Supreme Court. I included two typically blue states and two typically red states in my analysis, which may contribute to this potential political polarization between people from different states.

I chose to see if respondents and petitioners of cases presented to the Supreme Court were from the same or different states. To do this, I identified states using their variable representations in the codebook, I used case_when() to identify when states were the same or different, I grouped states by if they were the same state, I found the number of total cases, and I used a bar chart to represent the differences in the number of total cases between different states and same states. 

## Session info

```{r session-info}
sessioninfo::session_info()
```


## Reflections & Resources

Write 1-2 paragraph reflecting on what was difficult and/or easy about this homework assignment. Discuss the problems you encountered and how you solved them, and new concepts or techniques you learned.

Please list the first and last names of any collaborators you worked with to complete this assignment. 

Additionally, list the resources you used and provide the links, including how you utilized them (if you only used in-class resources, just say so, you do not need to list them). 

Kindly remember that you are not allowed to use AI tools to generate R code for this and future assignments. The only acceptable uses of AI tools are for debugging (but only after you have made an attempt on your own) and for generating examples of how to use a specific function (but also check the function documentation and course materials). If there are different ways to code something, you should always prioritize what learned in the course.

### Reflection
This was a very difficult homework assignment and it took me a much longer time than the previous assignments. Recoding variables took me a minute to get used to, but I got the hang of it quickly throughout Task 1. Task 2 I somewhat enjoyed because I could use the variables I just recoded in Task 1 and see the effects of the recoding on the data as a whole. After a few nonsense coding attempts at the make-your-own question, I really enjoyed coming up with that idea by myself. Task 3, however, was much harder. It took me a lot longer to figure out the logic behind what each part of each question was asking me to do, and I needed more help from ChatGPT to help me debug. It took me awhile to come up with my own question, but it ended up being easier to answer my own question than to answer the previous questions as I feel like I understood it better because I came up with it.

### Resources

- I primarily used in-class resources, and found the Ed discussion board particularly helpful because other students were asking the same questions I had.
- I used the `ggplot2` documentation to help me figure out the scale_y_continuous code.
- I used ChatGPT to help me debug when I encountered an error when I used geom_bar() for Task 3 Question 2. ChatGPT suggested I use geom_col(). I also asked it how to invert the coordinates of the graph without getting an error and it suggested I use coord_flip().